# Discussion
```{r vcsetup, echo=FALSE}

blades <- datasetlp %>% 
  filter(BlankType == "ElongatedProd" & MedWidth >= 12 & Phase == "Proto-Solutrean") %>% 
  select(Length, MedWidth, Thickness, BlankShape, PlatformThickness, PlatformWidth, OtherNotes, RawMaterial) %>% 
  mutate(Elongation = Length/MedWidth, Flattening = MedWidth/Thickness, PlatFlattening = PlatformWidth/PlatformThickness)

measur <- blades %>% 
  dplyr::summarise(Elongmean = mean(Elongation), ElongSD = sd(Elongation),
            Flattmean = mean(Flattening), FlattSD = sd(Flattening),
            PlatFlatmean = mean(PlatFlattening), PlatFlatSD = sd(PlatFlattening),
            n = n())

shape <- blades %>% 
  group_by(BlankShape) %>% 
  dplyr::summarise(n = n()) %>% 
  ungroup() %>% 
  mutate(freq = n/sum(n)*100)


VC <- dataset %>% 
  filter(OtherNotes == "TVC?") %>% 
  select(Length, MedWidth, Thickness, BlankShape, PlatformThickness, PlatformWidth, OtherNotes) %>% 
  mutate(Elongation = Length/MedWidth, Flattening = MedWidth/Thickness, PlatFlattening = PlatformWidth/PlatformThickness)


VCmeasur <- VC %>% 
  dplyr::summarise(Elongmean = mean(Elongation), ElongSD = sd(Elongation),
            Flattmean = mean(Flattening), FlattSD = sd(Flattening),
            PlatFlatmean = mean(PlatFlattening), PlatFlatSD = sd(PlatFlattening),
            n = n())

VCshape <- VC %>% 
  group_by(BlankShape) %>% 
  dplyr::summarise(n = n()) %>% 
  ungroup() %>% 
  mutate(freq = n/sum(n)*100)



blades_meas <- read_csv("data/blade_VC_measurements.csv")

elong <- ggplot(blades_meas, aes(x=Assemblage, y=ElongMean, fill=Class)) + 
  geom_bar(stat="identity", color="black", 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=ElongMean-ElongSD, ymax=ElongMean+ElongSD), width=.2,
                position=position_dodge(.9))+
  ggpubr::color_palette("jco")+
  theme_pubclean() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))


flat <- ggplot(blades_meas, aes(x=Assemblage, y=FlatMean, fill=Class)) + 
  geom_bar(stat="identity", color="black", 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=FlatMean-FlatSD, ymax=FlatMean+FlatSD), width=.2,
                position=position_dodge(.9))+
  ggpubr::color_palette("jco")+
  theme_pubclean() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))


plat_flat <- blades_meas %>% 
  filter(!Assemblage %in% c("Casal do Cepo", "Vale Almoinha"))

plat_flat <- ggplot(plat_flat, aes(x=Assemblage, y=PflatMean, fill=Class)) + 
  geom_bar(stat="identity", color="black", 
           position=position_dodge()) +
  geom_errorbar(aes(ymin=PflatMean-PflatSD, ymax=PflatMean+PflatSD), width=.2,
                position=position_dodge(.9))+
  ggpubr::color_palette("jco")+
  theme_pubclean() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))

converg <- blades_meas %>% 
  filter(Class != "Points a face plan")

converge <- ggplot(converg, aes(Assemblage, ConvergentEdges)) +
  geom_linerange(
    aes(x = Assemblage, ymin = 0, ymax = ConvergentEdges), 
    color = "lightgray", size = 4
  ) +
  geom_point(aes(color = Class), size = 4)+
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggpubr::color_palette("jco")+
  theme_pubclean() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))
```
The primary goal of this study was to understand the technological organization changes happening at the onset of the LGM, when the climatic conditions became abruptly harsh. To do so, we aimed to characterize two Proto-Solutrean/Terminal Gravettian assemblages at a technological level, understanding site specific patterns, then expanding these to what was already known for the Portuguese Estremadura, as a way of testing the existing models.

## Technology
Regarding Vale Boi, comparing its two identified groups, it seems obvious that there aren’t wide technological differences between the phases, both characterized by a high internal variability. In both groups, Lower 5 and Upper 5/4E, we see the presence of reduction sequences for the obtention of flakes, many of which are highly elongated, blades and bladelets, although Upper5/4E shows a larger number of blades. These are obtained through single platform cores, some of which are prismatic although there is also the presence of pyramidal, rarely with prepared platforms, with the occurrence of some faceted platforms in group Upper 5/4E, which might represent a further diversification of platform preparation. However, the tendency does seem to be for plain platforms or, mostly in quartz, fully cortical platforms. Quartz cores, in both groups, show the tendency for simple platforms cores, as well as smaller products, comparatively, for example, to chert, representing perhaps raw material constraints imposed only by the natural breakage patterns of the mineral. Blanks seem to be obtained mostly through simple unidirectional strategies, although bidirectional strategies are also employed at a smaller scale.

In both groups, there seems to be present several moments of the reduction sequence, with the removal of completely cortical blanks, progressively diminishing in percentages, although most of the assemblages had 0% of cortex, as is expected. Although in most cases, cortex type for chert wasn’t able to be determined, it is still visible the existence of a larger percentage of cortex in this raw material than others, comparatively, for example, with quartz and greywacke. Cores in this raw material seem to have been made on cobbles and pebbles, which were most likely collected in proximity to the site. Despite the low frequencies of cortex on quartz blanks, however, the high frequencies of cortical platforms may show the application of simple reduction strategies on unprepared cores on cobbles at the site.

The obtained products show a large frequency of convergent and parallel shapes, triangular cross sections and straight or curved profiles. This seems to be true for both flakes and elongated blanks, which, combined with the values for core elongation, may show the focus on a reduction strategy specifically for higher elongations blanks, many of which don’t fit the traditional definition of elongated blank, be it bladelet or blade. This seems to be further relevant in group Upper 5/4E which shows general higher elongation values for blanks in chert. However, in most all raw materials and groups, there are always high percentages of irregular blanks, which, in combination with the existence of several inform or simple platform cores might suggest the existence of an expedient reduction sequence.

The retouched tools  in Lower 5 and Upper 5/4E seem to show some differences in terms of quantity and variety, although this is to be expected, since it may reflect only the intensify of the occupation of Upper 5/4E, either representing differences in site occupation time or functionality (Cascalheira and Bicho 2013). One of the most obvious differences between the two groups is the presence of Vale Comprido points, limited to Upper 5/4E.

Frequencies of burins, either dihedral burins or burins on truncation, retouched tools relatively frequent during the Gravettian, but which seem to disappear during the Solutrean (Cascalheira and Bicho 2013; Cascalheira et al 2012), seem to remain stable and relatively similar through the two identified groups. 
	
Lapa do Picareiro is slightly harder to understand regarding reduction strategies, since the assemblage is small and truncated, not showing the complete sequence, a possible result of its functionality and high altitude, with difficult access. As has been discussed, the spatial dispersion of the materials, the raw material percentage differences and different dates have allowed the interpretation of two distinct occupations. These occupations show some differences, although, once again, as was the case with Vale Boi, their technological differences aren’t very obvious, although this is further complicated at Lapa do Picareiro due to the small truncated sample.

The small sample of cores shows the same types identified in Vale Boi, with simple platform, prismatic and pyramidal in the Middle T group, also marked by plain platforms and unidirectional removals. The flakes in the assemblage show similar patterns in between groups, such as parallel and convergent shapes, triangular or trapezoidal cross sections, straight profiles, unidirectional dorsal patterns and plain platforms. However, Middle T shows, compared to U/Lower T, a bigger prominence of convergent shapes in chert, higher frequencies of dihedral platforms, as well as higher flattening ratios. Elongated blanks show a higher level of attribute similarity between groups, with parallel and convergent shapes as well as straight and curved profiles, not differing highly from the flakes.

From the analysed cores and blanks, earlier moments of reduction sequence seem to be absent from the assemblage, for both groups. The absence of fully cortical blanks, as well as the general absence of debitage waste comparatively to other classes, but the presence of cores, may be understood as the transportation of almost completely non-cortical blocks of raw material for knapping at the site, or the transportation of finished blanks and tools, although the latter aren’t well represented.
One of the most obvious differences between the two groups within Lapa do Picareiro is the raw material usage, with levels U/Lower T being characterized by a high use of quartz, which shows similar products and production strategies to those used for chert, and Middle T, which shows a drastic inversion of quartz/chert use, with the latter being more frequent. One difference that can be observed in both groups, however, is the presence of chert blades, while quartz elongated blanks seem to be used mostly for the production of smaller elongated products, traditionally classified as bladelets.

Given the similar dates and raw material patterns, we may understand the assemblage from level U/Lower T (Lapa do Picareiro) and Lower 5 (Vale Boi) as Terminal Gravettian. Although the assemblages are complicated to compare, not only for the difference in sample dimension, but also due to the difference in the functionality and geographic location of the site, they seem to share many technological characteristics, such as the use of similar reduction strategies for quartz and chert (although this is mostly inferred for Lapa do Picareiro through a small sample of cores). For both these layers there is large quantity of quartz blanks, with higher frequencies than chert. From these, the elongated blanks in quartz, however, seem to be used preferentially for bladelets, whereas chert appears in the shape of bladelets and blades, although chert blades frequency seems fairly small for Lower 5 at Vale Boi. Similarly, both levels at these sites seem to show little platform preparation and the use of unidirectional strategies.

This use of quartz, as has also been noted for other sites with Terminal Gravettian assemblages in the Estremadura, is not a direct response to the absence of good quality chert (Almeida 2000). In Vale Boi, chert continues to be used primarily in blank and retouched tool production, while Lapa do Picareiro, belonging to the Estremaduran area, is in proximity of chert sources whose visibility patterns didn’t seem to have changed during the HE 2 (Zilhão 1997) with sites with Terminal Gravettian assemblages also using good quality chert as a primary raw material (Almeida 2000). Quartz, however, in the Terminal Gravettian sites was used for the intensive production of bladelets. In Vale Boi, for example, although there is an intense concentration of bladelets in quartz, more so than there are blades, the blank sample is dominated by flakes. This follows the patterns observed in Gato Preto, where the exploited quartz was of coarse and medium quality, and the main types of blanks are flakes (Almeida 2000). A similar pattern may be present in Vale Boi, where the assemblage is dominated by coarse quality quartz, which appears mostly in the shape of debitage waste, while blanks were mostly produced in fine and medium quality quartz, which is probably related to the inherent knapping properties of the different types of quartz. Lapa do Picareiro seems to show exactly that preference, where coarse and medium quality quartz is barely present, and most blanks are produced in fine quality quartz and rock crystal. This last type of quartz shows, however, a deviation from the dominant patterns in the Terminal Gravettian Estremaduran sites, where rock crystal were rarely used (Almeida), although this may be a problem of raw material availability.

These patterns allow us to understand then, that for Lower 5 and levels U/Lower T, quartz was used as a primary raw material, which was not justified by the lack of good quality chert, and with similar reduction sequences as chert, which was also used in high proportions.

Almeida (2000), still referring to raw materials, suggests the existence of workshops during the Terminal Gravettian, although none have been found, based on the transportation of chert into the sites in the shape of preformed cores, unretouched blanks and tools. Given the characteristics of both Vale Boi, which shows a high frequency of debitage waste and products, and thus had knapping happening at the site, and Lapa do Picareiro, which doesn’t show signs of intensive occupations, it isn’t possible in the present study to corroborate, or not, the suggestion. However, the presence of dolerite at Vale Boi, in the shape of blanks and tools, without the obvious presence of cores or debitage waste, may hint that such workshops existed in the Algarve, although this will be further discussed below.

The patterns already recognized regarding cores and platform preparation for Lower 5 and U/Lower T also seem to be in concordance with data from Terminal Gravettian assemblages, where platforms are mostly plain and there is little existence of crest preforming for the obtention of blades (Almeida 2000). Likewise, Almeida (2000) shows that for quartz, there is also the presence of cortical platforms, and especially in Gato Preto, reduction sequences were simple. This seems to be the case for Vale Boi as well, where not only are platforms plain but also cortical, with simple single platform cores as well. This repeated similarity with Gato Preto may, once again, be related to quartz quality.

Regarding elongated blanks, data from Estremadura shows a further relevant production of bladelets, in comparison to blades (Almeida 2000). Although there is a relevant number of blades at Lower 5 in Vale Boi, there seems to be an obvious larger amount of bladelets comparatively, which may be understood as a focus on bladelet/smaller product production rather than larger products. This may, however, only reflect selective transportation of blades outside of the site, since chert does show some gaps in elongated blank width and length dispersion, which doesn’t seem to be related to the existence of different reduction sequences for the production of blade and bladelet. Still, given the wide amount of bladelets in Lower 5, it seems that Vale Boi follows the patterns described for Estremadura. Although production may not be inferred through Lapa do Picareiro results, it seems obvious that there are more bladelets present in U/Lower T, although in this case, the presence of blades is also very relevant. This seems to follow the patterns identified for the Terminal Gravettian assemblage of Cabeço de Porto Marinho, where the production of blades was significant (Almeida 2000), which shows some variability in size of preferred blank production within the Terminal Gravettian, although, once again, at Lapa do Picareiro, human selection and transport are the main behaviours which can be inferred.

Alike the Estremaduran assemblages, bladelets in both Lower 5 and U/Lower T are rarely retouched (Almeida 2000), although, as mentioned in Chapter 7, in regards to Lapa do Picareiro this may be related to retouched identification issues. Despite the low frequencies of retouched bladelets that appear on Vale Boi, comparatively to the results shown for Estremadura, there is the lack of carinated endscrapers, elements which have characterized the Terminal Gravettian, giving it similarities with the Aurignacian.

Almeida (2000) has also noted the absence of Vale Comprido reduction strategies in Terminal Gravettian sites from Estremadura, something that seems to happen as well in Lapa do Picareiro and Vale Boi. This is further obvious when, in the case of Vale Boi, there is a clear concentration of Vale Comprido points in levels from Upper 5 and 4E, whereas Lower 5 shows no points, as well as Vale Comprido blanks.

As such, both groups Lower 5, from Vale Boi, and U/Lower T from Lapa do Picareiro, show technological and raw material preferences in concordance with those identified in Terminal Gravettian assemblages from the Estremadura: high percentages of quartz exploitation, though similar reduction strategies as chert; exploitation characterized by flake and bladelet production; reduced platform preparation. If for Lapa do Picareiro, the attribution of U/Lower T to the Terminal Gravettian had already been made, for Vale Boi, it may seem that what had been previously considered a single Proto-Solutrean occupation, does, in fact, reveal an initial moment of occupation, in Lower 5, which corresponds to a Terminal Gravettian culture, at around 26 kcal BP, and, despite the geographic distance to Estremadura, shows similar patterns to those in the Estremadura.

Comparatively to the Terminal Gravettian, the Proto-Solutrean shows a diminishing use of quartz in the transitional model established through the technological analysis of sites within the Estremadura, although reduction strategies similar to chert continue to be used (Zilhão 1997). This pattern seems to be the case for level U/Lower T of Lapa do Picareiro (which may be the result of transport and not obvious production preferences) and Upper 5/4E in Vale Boi, although, in the latter, quartz percentages continue to be fairly high, at one point dropping to nearly 30% but never becoming as high as Lower 5.

Results from the Estremadura have shown, for the Proto-Solutrean, the existence of three different operative sequences, for the production of blades, bladelets and Vale Comprido blanks. According to Zilhão (1997), however, the existence of a separate blade reduction sequence may be absent in the fully developed Proto-Solutrean. This seems to be the case for Upper 5/4E, where the production of elongated blanks, just as was the case from Lower 5, doesn’t seem to show a marked bimodal curve, in chert for example, which may reflect the continuous reduction of core volume resulting in the production of gradually smaller products.

Likewise, the production of smaller blanks through carinated endscrapers and carinated elements, which is present in the Terminal Gravettian, seems to be absent in the Proto-Solutrean of the Estremadura (Zilhão 1997). However, Upper 5/4E of Vale Boi shows the presence of these tools/cores (unlike the Lower 5 group) even if in rather small percentages, although this might only be representative of the assemblage’s wide variability regarding tool types. Even so, it seems that, at least for Vale Boi, “Aurignacian V” carinated elements are still present.

One of the marked differences between the Terminal Gravettian assemblages and Proto-Solutrean ones may be the focus on obtained products size, where in the first, as beforementioned, there is a wide production of bladelets with smaller numbers of blades, while in the Proto-Solutrean, perhaps with the introduction of a Vale Comprido production sequence, blade frequency becomes increasingly relevant. This can be seen in Upper 5/4E of Vale Boi, where frequencies of blade-sized elongated products stands with more relevance than Lower 5, most of these being in chert. For Lapa do Picareiro however, in Middle T, these values aren’t as obvious, since the maintain the same relevance as in group U/Lower T, even if, as noted, this may be the result of selection and transportation issues, and not represent actual production patterns.

Aside from the production of blades and bladelets through prismatic cores, which seem to maintain mostly the technological patterns seen in the Terminal Gravettian, the Proto-Solutrean technological innovation stems from the introduction of a Vale Comprido reduction sequence, which is characterized for cores with pyramidal tendencies (which may be understood as higher convergency values), with unprepared platforms, from which obtained products are elongated blanks (either flakes or blades) with convergent shapes and thick platforms. This reduction sequence is eventually directed towards the production of Vale Comprido points, which mark one of the retouched tool innovations of Upper 5/4E of Vale Boi. In fact, Vale Boi not only has a concentration of Vale Comprido points in this group, as a large group of similar blanks (flakes and blades) with the above mentioned characteristics, which have allowed for the identification not only of Vale Comprido points, as well as possible Vale Comprido technology, in the form of unretouched blanks. It is the presence of this type of reduction sequence and technology that, in conjunction to other changes in technology, allow for the attribution of Upper 5/4E to a Proto-Solutrean occupation.

Despite the other technological similarities of Middle T’s (Lapa do Picareiro) assemblage with both Upper 5/4E of Vale Boi and the results from Estremadura, Vale Comprido technology doesn’t seem to be so obviously present in site.  Both the retouched tool and blanks identified as product of this reduction strategy show similar attributes to those of Vale Boi and other sites of Estremadura. Platforms seem generally plain and obtained products are elongated and convergent, with lengths in between 50-70 mm, only the platforms looking rather different. In fact, one of the noticeable characteristics of Vale Comprido technology, more so in the blanks since these have not yet been thinned through retouch at the platform, is the thickness of their platform and general blank. On Lapa do Picareiro, in Middle T however, blanks possibly resulting from Vale Comprido-like reduction sequences, show thin cross sections and platforms, resulting in similar, but different looking products which cannot be, at first look, classified as Vale Comprido technology.


## Vale Comprido technology

If the attribution of a Proto-Solutrean occupation to Upper 5/4E in Vale Boi can be easily made through both technological characteristics and the presence of a Vale Comprido reduction sequence, the same seems to be further difficult for Middle T at Lapa do Picareiro. Since the identification and comparison of Vale Comprido technology, however, has been made through morphological data and ratios, the present study applied the same methodology, comparing the values for Vale Boi and Lapa do Picareiro’s Vale Comprido blanks and points, in order to understand the actual differences between this technology, as well as comparing it to blanks and points of Proto-Solutrean assemblages, such as Vale Comprido (Encosta) and of other chronologies, such as Casal do Felipe (Aurignacian), Vale de Porcos (Gravettian), Casal do Cepo and Vale Almoinha (Middle Solutrean).

Elongation means (\ref{fig: plotsVCelong) show through all of the groups somewhat similar ratios of elongation, with the lowest values appearing in Vale Boi, Vale Comprido – encosta, for the points, Lapa do Picareiro, Casal do Cepo and Vale Almoinha for the points à face plan, although the differences aren’t very significant.

However, the lower elongation values for blades and points from these sites, especially from those with Proto-Solutrean assemblages, might reflect the use of both blades and elongated flakes for the production of the points, thus having a reduction sequence focused on the production of elongated blanks, without the goal of obtaining products with high elongation ratios.

```{r plotsVCelong, echo=FALSE}
plot(elong)
```

Regarding flattening (mesial width / thickness, \ref{fig: plotVCflat}) there seems to be a group of sites with smaller values, and thus with blades and points which are less flat: Vale Boi, Vale Comprido – encosta and Vale de Porcos. This seems to be in concordance with the already established patterns for Proto-Solutrean assemblages, where the production strategies seem to focus thick blanks which are then thinned at the platform. However, it is possible to notice the association between the Lapa do Picareiro blades with other assemblages, with blades and points with higher levels of flattening, thus showing that the Vale Comprido blades show some morphological differences in the flattening ratios from Vale Boi and Vale Comprido – encosta, both with Proto-Solutrean assemblages.

```{r plotVCflat, echo=FALSE}
plot(flat)
```

A similar pattern can be observed for platform flattening means (\ref{fig:plotVCplatflat}), although the differences between Lapa do Picareiro and Vale Boi or Vale Comprido – Encosta don’t seem as obvious. Using this morphological indicator, Casal do Felipe shows the widest differences, with high values for platform flatterning, while the other groups seems rather similar. It is possible to observe, however, that Lapa do Picareiro shows a wide dispersion and variability in terms of platform flattening, comparing, for example, to the low values and low standard deviations from Lapa do Picareiro and Vale Comprido – Encosta.

```{r plotVCplatflat, echo=FALSE}
plot(plat_flat)
```

Regarding the convergence of the edges, figure \ref{fig:plotVCconverge} shows a distinct difference in the percentage of convergence between Lapa do Picareiro, Vale Boi (blades and points) and Vale Comprido – encosta (blades and points). The blades from these sites show the highest percentages of convergent edges (above 30%) from all the groups, only surpassed by the values from Vale Comprido points, which reach 50% in Vale Comprido – Encosta and more than 60% in Vale Boi. All the other blades from sites with assemblages belonging to the Solutrean, Gravettian or Aurignacian show values of 20% or under. This shows not just the already established preference for convergent edges in Vale Comprido points, as well as a high component of convergence in the blank assemblages of these three sites, more specifically, in the blades.

```{r plotVCconverge, echo=FALSE}
plot(converge)
```

As such, it seems that the analyzed morphological indicators are good for understanding some variability between the groups and assemblages, as well as distinguishing between blades and points from the same site, which might be useful to understand blade production within the Proto-Solutrean assemblages with the goal of obtaining Vale Comprido points. Although, as seen, all assemblages seem to have a wide variability regarding morphological attributes, which creates difficulties in understanding specific groups based on the means of those indicators. This is especially noticeable with the elongation ratio mean, which didn’t show very evident differences between the blade assemblages from the different technocomplexes, although this might also be related to a certain homogeneity in the elongation of blades throughout the Upper Paleolithic.

Regarding the ratios and morphological attribute analysis, Vale Boi seemed to be in concordance with the Vale Comprido – Encosta values, for both the points and the blanks. The results for the blanks also showed wide similarities with the points, indicating the production of blades is related to the production of the Vale Comprido points. 

However, from all of the attributes analyzed, it was the frequency of convergent edges that showed the biggest differences between the Proto-Solutrean assemblages and Lapa do Picareiro against the other assemblages with lesser percentages of convergent edges. This clearly shows that convergence is an important attribute to define the assemblage, its blanks and points.

Despite this similarity, Lapa do Picareiro, especially within the flattening ratios, showed values which were closer to those from the Solutrean assemblages, with flatter blades than those from Vale Boi and Vale Comprido – Encosta.

These differences, along with the difficulty in identifying Vale Comprido points in the site’s Middle T assemblage, for the similarities with the type but lack of defining qualities such as the thickness of the platform or of the blank, and the fairly late chronology attributed to Middle T in Lapa do Picareiro, might represent the nonexistence of a Proto-Solutrean assemblage at the site. Instead, and given the similarities of the Lapa do Picareiro blades with both Proto-Solutrean and Solutrean blades, the assemblage of Middle T might be better explained by a transitional assemblage, chronologically located somewhere in between the Proto-Solutrean and the Middle Solutrean.

## Chronology
As seen on Chapter 2, the Proto-Solutrean has been explained as a transitional phase model developed with basis on the technological characteristics of specific assemblages in the Portuguese Estremadura, each with chronological limits, also set by radiocarbon dates from these sites. When comparing those dates with those obtained in Vale Boi and Lapa do Picareiro, it seems apparent that what looked like a streamlined range for the Estremadura results (consequence of the nonexistence of date for sites with Vale Comprido points), is, in fact, for Vale Boi and Lapa do Picareiro, two groups of dates: an older group, between the 26.6 kcal BP to 26 kcal BP (at the lower boundaries) and a second group of younger dates, starting with Vale Boi at 24.7 kcal BP (lowest boundary). All the Estremaduran dates either overlap with the group of older dates or fits in between the two groups.

Observing this information, it seems obvious that there is a well defined moment shared by the three sets of dates (Vale Boi, Lapa do Picareiro and Estremadura after Zilhão 1997). This moment, setting itself around the 26.4 kcal BP is accompanied, in all sites, with changes in raw material use, some technological differences and typological innovations, reflected at Vale Boi in the Lower 5 occupation (less dense) and at Lapa do Picareiro by layer U/lower T occupation. As such, it seems this moment, as already discussed for the Estremaduran sites, represents a Terminal Gravettian occupation, which seems to occur at similar chronologies even in the south of Portugal. 

The Proto-Solutrean phase has been theorized to having occurred around 25.5 kcal BP and 25 kcal BP. The group of dates from Estremadura and Vale Boi seem to reflect just this, with changes in raw material percentages and technological differences occurring in the Upper 5/4E group in Vale Boi, as well as the appearance of Vale Comprido points as abovementioned. Using both the chronology and technological patterns, we can, thus, identify the Upper 5/4E group as a Proto-Solutrean occupation, possibly occurring simultaneously with the Estremaduran occupations, although perhaps lasting a bit longer, although this might only represent the lack of radiocarbon dates for Proto-Solutrean contexts in the Estremadura (as discussed on Chapter 2).

Although it would be expected that Vale Boi either showed different chronologies and technological patterns from those observed in other sites, since it appears isolated from the cluster at Estremadura due to the geographic distance, as has been noted in the Gravettian horizons (Marreiros 2009), instead, what we find is a high level of similarity. This seems to corroborate the idea that the Proto-Solutrean was a moment of expansion of communication networks, thus leading to a higher degree of homogeneity within assemblages throughout the territory (Cascalheira and Bicho 2013) which occurred simultaneously in the territory. This seems to be the case for other sites outside of the Iberian Peninsula as well, where Proto-Solutrean assemblages have been recognized with similar techno-typological characteristics as those identified in the Estremadura (Renard 2011). Once again, understanding the Proto-Solutrean as a moment of cultural breakage onset by abrupt harsh climatic conditions, the expansion of these networks and spread of the culture through the territory may have functioned a new adaptative behavior, perhaps also permitted by the changing conditions of the vegetation, since, as seen on Chapter 2, the HE 2 is marked by receding Mediterranean adapted forests, substituted by wider desertic biome shrubs, which may have facilitated mobility through the territory.

At Lapa do Picareiro, alongside the technological differences and nonexistence of a true Vale Comprido technology, there is also the association of Middle T with dates of around 24 kcal BP. These dates, instead of overlapping with those of Vale Boi Upper 5/4E (which would somehow be expected if both assemblages represented Proto-Solutrean occupations), show two interesting features: 1) they are far apart from those of levels U/Lower T, showing there are, in fact, two different occupations, already visible through the spatial and raw material dispersion, now corroborated by chronological differences, showing in between a moment of occupational hiatus ; 2) the group identified as Middle T doesn’t seem to have occurred simultaneously as the Proto-Solutrean occupation at Vale Boi, showing instead younger dates which place these levels as much more recent than those expected in the Proto-Solutrean transitional model, and before the dates for the Middle Solutrean.

As such, we may understand the previously referred techno-typological and chronological characteristics of Middle T of Lapa do Picareiro, not as a Proto-Solutrean occupation, which, according to this interpretation, didn’t seem to happen, characterized only by a hiatus of occupation of the cave, but perhaps as a succeeding moment of transition, characterized by flatter/thinner blanks and technology, alike those to be expected in Middle Solutrean technology, but that also resembles Vale Comprido technology in terms of blank morphology and shapes, connecting the Proto-Solutrean to the Middle Solutrean, a stage that in France has been understood as the Lower Solutrean, but nonexistent in Portugal until the present moment (Zilhão 1997).

## Raw material use and preference
Aspects as the high exploitation of quartz, and application of similar reduction strategies to both chert and quartz, can be integrated in a perspective of cultural diversification and techno-economical intensification. These patterns have been interpreted as preferences resulting from tradition, more than technical or economic reasons (Almeida 2006). This interpretation steams from the understanding that, for the Estremaduran area, there is not a lack of good quality chert (Zilhão 1997) since, even during times when quartz was highly used, chert continued to frequently appear in the archaeological record, while there seemed to be a preference for this raw material for formal tool production (Almeida 2006). The same seems to apply for Vale Boi, where the collection in either phases shows the existence of good quality chert, in fact, most of retouched pieces being produced in this raw material, including Vale Comprido points.

However, when considering the Proto-Solutrean at the onset of a harsh climate event and accompanying environmental changes, these patterns and preferences may be better explained by growing pressures of the cultural system, and the ability for such a system to cope by expanding the economical options, and thus using a previously secondary raw material, which was closely present, even if in differing qualities, as a primary one, along aside the already used raw material, when this pattern cannot be better explained by raw material visibility, onset by environmental changes, which seems to be the case (Cascalheira and Bicho 2013). As such, the shifts in quartz use might be explained by the climatic shifts of HE2 and at the onset of the LGM, and their impacts on ecology throughout the Iberian Peninsula, even if climate changes weren’t as harsh as other areas in the European continent.

Understanding the Proto-Solutrean as a Three-Phase model, and thus separating the Terminal Gravettian from the Proto-Solutrean, also separating quartz use from very frequent to less frequent, forces us to admit that this technological diversification happened only in a “short” moment before the Solutrean. This, although, may be explained through the Panarchy framework (Holling 2001) and RRM (Bradmöller 2010), where moments of higher climatic instability provoke system breakdowns, which, in this case, may be seen through the different raw material use and procurement choices, which correspond to a moment of rapid reorganization, recombination which leads to innovations. These innovations, however, are tested, used and while some fail, and disappear, others survive and adapt, starting once again the loop. As such, the rapid climatic changes brought by the HE2 at the onset of LGM may have triggered weaknesses in the previous economic system, forcing the hunter-gatherer communities to expand their technology to the use of raw materials such as quartz, creating thus an innovative response, which was highly used during the Terminal Gravettian, following a progressive reduction during the Proto-Solutrean, to its failure and adaptation of raw material systems during the Solutrean, where quartz is, once again a secondary raw material, initiating a new cycle.

The same model may be applied to another raw material in Vale Boi’s archaeological record which only appears in the Proto-Solutrean levels, the dolerite. Thus, using the panarchy framework, dolerite may be been an innovative response in the upper occupational phase to system breakdowns onset by climatic pressures and ecological changes, but which failed, thus disappearing from the archaeological record during the following Middle Solutrean levels. However, the failure of this raw material may be related to the failure of a technology for which it was preferentially used, as are the Vale Comprido points, which fits equally in the model, and during the same phase.

There is, however, a caveat in this interpretation. As mentioned before, dolerite was identified based on its texture and coloration, which visibly differed from all raw materials in the assemblage and site. Through its unaltered characteristics, however, on the second moment of analysis, a few other pieces in the same raw material, and without the reddish patina, were identified. From here, the present study may formulate two hypothesis regarding the presence of dolerite in Vale Boi: 1) H0, which says dolerite use can only be observed in the Proto-Solutrean levels, with or without patina; 2) H1, which says dolerite use, probably without patina, may be observed in other archaeological levels. 

Hypothesis one is based on the fact that unpatinated dolerite closely resembles fine grained sandstone, even at a mineral level, and thus, may have been considered as sandstone, greywacke or another raw material during previous analyses, which may only be tested through the analysis of Vale Boi raw materials through all layers, with the combination of physical and chemical approaches.